When Forgiveness Backfires: Rejection Sensitivity and Cooperative Behavior Following Exposure to Adaptive Forgiving Agents

### 2025-12-05 article.Rmd
Revise manuscript: add event study analysis, strengthen limitations, update terminology
- Add event study and pre-low-investment analyses to test contrast vs betrayal aversion
- Acknowledge confound between forgiveness and low investment exposure in Limitations
- Change "defection" to "low investment" terminology throughout
- Reframe RS findings: intact detection but impaired recovery in high RS
- Condense Player Ratings section; revise abstract and discussion

---
title: "When Forgiveness Backfires: Rejection Sensitivity and Cooperative Behavior Following Exposure to Adaptive Forgiving Agents"
author:
  "Ismail Guennouni^1,2,3,4^*, Georgia Koppe^1,2,3^$^\\dagger$, Christoph Korn^4^$^\\dagger$"
abstract: |
bibliography: "bib/exposure.bib"
csl: apa.csl
output:
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
header-includes:
  - \usepackage{lscape}
editor_options: 
  markdown: 
    wrap: 72
---

@@

Can exposure to forgiving behavior improve cooperative outcomes, and
does Rejection Sensitivity (RS) moderate this effect? This randomized
controlled experiment explores these questions using a novel approach
with Hidden Markov Model (HMM)-based artificial agents in repeated trust
games (RTGs). Participants (N = 206), pre-screened for high or low RS,
interacted with either consistently behaving or more forgiving
co-players, which were in reality HMM agents. Unexpectedly, exposure to
forgiving agents led to reduced cooperation in subsequent interactions.
Control group participants maintained consistent behavior. RS levels
influenced perceptions of the agent's cooperativeness but did not
significantly affect behavioral outcomes, revealing a
perception-behavior dissociation. These findings challenge assumptions
about fostering cooperation through simple exposure to positive
interactions and highlight the importance of accounting for negative
contrast effects when designing such interventions. They also
demonstrate the potential of HMM-based artificial agents for studying
interactive social dynamics, offering a methodological advancement for
future research. Our results suggest the need for nuanced interventions
that gradually shape expectations and behaviors to foster cooperative
outcomes, particularly addressing the discrepancy between perceptions
and behavior in individuals with high rejection sensitivity.

@@

This study reveals that exposing individuals to forgiving behavior in
economic games can unexpectedly decrease their subsequent cooperation.
Interestingly, while people with high rejection sensitivity more
accurately perceived their partners' cooperativeness, this didn't
translate into more cooperative behavior. These findings challenge
assumptions about fostering cooperation and highlight the need for more
nuanced interventions to promote reciprocity, especially for individuals
prone to expecting rejection.

@@

@@

One explanation for such social challenges lies in early caregiver
experiences. Attachment theory [@bowlby_attachment_1978] suggests that
the quality of these relationships shapes our capacity for secure
attachments and trust. Individuals with higher levels of insecure
attachment may recall negative trust-related experiences more easily,
report fewer positive trust experiences, and use less constructive
coping strategies when trust is broken [@mikulincer_attachment_1998].
These insecure attachment patterns are often associated with heightened
rejection sensitivity (RS), a tendency to anxiously expect, readily
perceive, and intensely react to rejection [@downey_implications_1996;
@downey_early_1997]. RS has been linked to the development of various
mental health conditions, including depression, anxiety, personality
disorders, and self-harm [@gao_associations_2017]. Individuals high in
RS show attentional biases towards social threat cues, which may
contribute to difficulties in social interactions
[@berenson_rejection_2009]. A recent meta-analysis revealed prosocial
behavior and interpersonal trust as two key processes of interpersonal
functioning that are markedly impaired in PDs and which are likely to
contribute to interpersonal dysfunction in this population
[@hepp_prosociality_2022]. The interaction of RS and trust-based
constructs has been explored, particularly in Borderline Personality
Disorder (BPD). @miano_rejection_2013 and @richetin_emotional_2018 found
that RS mediated the relationship between BPD features and lower trust
appraisal. @abramov_influence_2022-1 found that higher baseline feelings
of rejection in individuals with BPD predict slower trust formation and
less pronounced declines in trust following trust violations during the
trust game. However, the interaction between *reciprocity* and RS hasn't
been studied as extensively, leaving a gap in our understanding of how
these constructs might interplay. Could an ingrained sensitivity to
rejection contribute to reduced pro-social behavior such as a failure to
show reciprocity as a result of perceived rejection by the interaction
partner?

Given that RS may be a manifestation of maladaptive attachment styles,
it's crucial to explore whether exposure to consistently forgiving
interaction partners could help reshape these interpersonal expectations
and behaviors. If RS indeed influences reciprocity in repeated
interactions, could such exposure serve as a corrective experience,
potentially improving interpersonal outcomes? This approach aligns with
attachment theory, which suggests that new, positive relational
experiences can modify internal working models of relationships
[@bowlby_secure_1988]. Research in the fields of behavioral economics
and psychology has also explored how positive social interactions
influence cooperation and pro-social behavior. The use of the repeated
trust game (RTG), a well-established experimental approach, has allowed
for the analysis of the development of trust through ongoing
interactions [@joyce_trust_1995]. In this paradigm, cycles of mutual
trust, where each party's trust is reciprocated with trustworthiness,
have the effect of enhancing cooperative behaviors even among
individuals who are initially inclined to be distrustful
[@king-casas_getting_2005]. @fowler_cooperative_2010 studied behavior in
social networks interacting in a public goods game and found that
cooperative behavior tends to cluster, suggesting that exposure to
cooperative peers can lead to more cooperative behavior. Similarly,
research on social learning theory [@bandura_social_1977] has long
demonstrated that individuals learn and model the behavior of those
around them, indicating that if someone is consistently exposed to
cooperative and positive individuals, they're likely to emulate this
behavior. These insights highlight that engaging with compassionate and
forgiving others can be an effective method for mitigating deeply
ingrained mistrust and potentially fostering trustworthy behavior.

@@

We pre-screened participants for high or low RS using a validated
questionnaire, then assigned them exclusively to the trustee role in a
series of trust games. After playing a 15-round RTG with a human-like
HMM investor, they were randomly assigned to either a Control or
Manipulation condition. In the Manipulation condition, participants were
exposed over three RTGs to HMM investors designed with a limited
propensity for retaliation, potentially mitigating ingrained mistrust.
In the Control condition, participants played three RTGs against the
same human-like HMM. After this exposure phase, all participants played
another 15-round RTG with a human-like HMM investor, similar to the one
in the pre-exposure phase. We expect those exhibiting high RS to be more
responsive to the pre-programmed change in the agent's cooperativeness
and retaliation propensity. We also hypothesize that those in the
Manipulation condition would behave more cooperatively and have a lower
propensity to retaliate to the co-player's defection after the exposure
phase, with those in the higher RS group showing a bigger effect.

@@

The experiment had a 2 (Condition: Manipulation or Control) by 2 (RS :
High or Low) by 2 (Phase: Trust-Game Pre Manipulation, Trust-Game Post
Manipulation) design, with repeated measures on the Phase factor (Figure
\@ref(fig:HMMPanels).A). Participants within each pre-screened RS group
were randomly assigned to one of the two levels of the Condition factor,
resulting in 101 participants in the Manipulation condition and 105 in
the Control condition. The games were designed and implemented online
using Empirica v1 [@almaatouq_empirica_2021]. The planned experiment
received approval from the University of Heidelberg's Medical Faculty
ethics commission (ID:S-708/2023) and the experiment was performed in
accordance with the ethics board guidelines and regulations. All
participants provided informed consent prior to their participation.

@@

In order to instigate a potential breakdown of trust, thereby allowing
us to probe efforts to repair it, the computerised agent was programmed
to provide a low investment on round 12 (pre-manipulation) and round 13
(post-manipulation). On all other rounds, the investor’s actions were
determined by randomly drawing an investment from the state-conditional
distribution, with the state over rounds determined by randomly drawing
the next state from the state-transition distribution as determined from
the net return on the previous round (disregarding the net return
immediately after the pre-programmed low investment rounds).

@@

At the start of the experiment, participants provided informed consent
and were instructed the study would consist of three phases in which
they would face a different other player. Participants were told their
goal was to maximise the number of points in all phases. They were not
told the number of rounds of each phase. Participants were randomly
assigned to either a Control or Manipulation condition. The timeline of
the experiment is shown in Figure \@ref(fig:HMMPanels).A. Phase one
("pre") consisted of a 15 round RTG in which participants took the role
of trustee, facing the same investor over all 15 rounds. On each round,
after being informed about the amount sent by the investor participants
decided how much of the tripled investment to return to the investor,
before continuing to the next round. Phase 2 ("exposure") consisted of
three 7-round RTGs. Participants in the Manipulation condition faced the
forgiving HMM investor and rated the agent on the same attributes as in
the pre-manipulation phase. Those in the Control condition faced the
same human-like HMM agent as in the "pre" phase and rated each co-player
on the same attributes. To keep the experience similar to the "pre"
phase, the agent in the Control condition was also designed to send a
very low investment in round 5 of each of the three games. In the
post-manipulation phase ("post"), participants in both conditions faced
the same human-like HMM as in "pre" phase.

@@

````{=tex}
\begin{landscape}
```{r HMMPanels, fig.cap= "\\small{A: A timeline of the experiment for both conditions. The RTG is played in dyads, with participants always assigned the role of the trustee and the HMM agent that of the investor. The investor is endowed with 20 units at the start of each round. They need to decide how much of that endowment they want to invest with the trustee. The investment is then multiplied by a factor of 3 and sent to the trustee who needs to decide how much of the multiplied investment they want to send back to the investor. The difference between conditions is the type of agents participants are exposed to in the exposure phase. Panels B - D: We construct the artificial investor agent by fitting a three-state hidden Markov model to data of human investors engaged in the 10 round RTG. From the fitted HMM, we get the distribution of investments by the human-like agent, conditional on its latent state as shown in Panel B. The fitted HMM also yields the transition probability of the agent to a state on trial t+1 as a function of the net return (difference between the investment sent and the amount received in return) on trial t as shown in Panel C. Each plot in Panel C represents a different starting latent state on trial t, and each line represents the probability of transitioning to a particular state in trial t+1. Panel D shows the transition probabilities of the forgiving HMM agent in the low trust state. Unlike the human-like HMM, the forgiving HMM always transitions out of the low-trust state, and is more likely to end up in a high-trust state, as a proxy for coaxing behavior. Transitions in the medium and high trust states were identical for both agents.}",include=T, out.width='1\\linewidth', out.height='1\\textheight', fig.align='center'}

@@

# Calculating the width and height for each image
width_per_image <- ((12 - 0.1) / 2) * dpi
height_per_image <- ((7.4 - 0.1) / 2) * dpi
# The geometry string for each image, including spacing (assuming no borders)
geometry_string <- paste0(width_per_image, 'x', height_per_image, '+', spacing, '+', spacing)

@@

To test whether participants behaved differently in the RTG after the
manipulation compared to the Control group, we model the percentage
return (percentage of tripled investment returned to investor) using a
linear mixed-effects model with Phase (RTG game pre vs.
post-manipulation), Condition (manipulation vs. Control), Investment,
and RS (High vs Low RS group) as well as their interactions as fixed
effects, and player-wise random intercepts and slopes for Phase. The
full specification of the statistical model can be found in the
supplement. To test whether participants behaved differently between
conditions in the Exposure phase, we fit a linear mixed effects model to
participants returns only in the exposure phase, with Condition
(manipulation vs. Control), Investment, and RS (High vs Low RS group) as
well as their interactions as fixed effects, and player-wise random
intercepts. Finally, to test whether the HMM agent's behavior differed
between Phases, Conditions and RS groups, we estimate a linear
mixed-effects model of investments sent by the computerised HMM agent
with Condition, Phase and RS and their interaction as fixed effects, and
a similar random effects structure to the returns model.

The model was estimated using the `afex` package [@singmann_afex_2022]
in R. More complex models with additional random effects could not be
estimated reliably, and as such the estimated model can be considered to
include the optimal random effects structure
[@matuschek_balancing_2017]. A similar process was used to establish the
random effects structures of linear mixed-effects models used to analyse
the HMM agent investments as well as the participants' ratings of the
co-players. There is no agreed upon way to calculate effect sizes for
mixed effects models. Instead, we will report on testing differences in
marginal means. For the $F$-tests, we used the Kenward-Roger
approximation to the degrees of freedom, as implemented in the R package
"afex". We Z-transform the Investment variable (subtract the overall
investment mean and divide by overall standard deviation) as centering
is beneficial to interpreting the main effects more easily in the
presence of interactions. To probe significant interactions, we
conducted planned contrasts using the `emmeans` package in R. Given that
we were testing multiple pre-planned comparisons, we applied the "Sidak"
correction to control for familywise error rate while maintaining
reasonable statistical power. This approach allowed us to investigate
specific hypotheses about the differential effects of our manipulation
across phases and RS groups, while protecting against inflated Type I
error rates.

@@

```{r}
# average investment for first 10 rounds pre manipulation
avg_pct <- avg_ret_df %>% filter(roundNum < 11 , gameNum.f == "pre") %>% summarise(avg_inv = mean(inv_pct), avg_ret = mean(ret_pct_na, na.rm=TRUE))
avg_pct
```

@@

```{r gamesPlot, include=T, echo=FALSE, fig.cap="Averages and standard errors of the trustee's return as a percentage of the multiplied investment received (y-axis) by Condition, Phase, and game round (x-axis) averaged across RS groups. The blue line shows the returns in the Pre phase and the green line those in the Post phase. The left Panel shows returns in the Control condition and the right one those in the Manipulation condition. The dotted lines identify the rounds where the pre-programmed one-off low investment occurs. We note lower average returns post vs pre in the Manipulation condition, whilst returns in the Control condition are similar between the two phases.",fig.align="center", fig.width=6, fig.height = 4}

# Create a data frame for the vertical lines
vline_data <- data.frame(
  xintercept = c(12, 13),
  Defection_round = factor(c("Pre-manipulation", "Post-manipulation"), levels = c("Pre-manipulation", "Post-manipulation"))
)

# Plot
ggplot(avg_ret_df %>% filter(gameNum.f %in% c("pre","post")), aes(x=as.factor(roundNum), y=ret_pct_na, group=gameNum.f, color = gameNum.f, fill=gameNum.f)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun.data = mean_se, geom = "ribbon", aes(ymin=..ymin.., ymax=..ymax..),
               alpha = 0.3, linetype = 0) +
  geom_vline(data = vline_data, aes(xintercept = xintercept, linetype = Defection_round), color = "black", linewidth = 0.6) +
  scale_linetype_manual(values = c("Pre-manipulation" = "dotted", "Post-manipulation" = "dashed"),name = "Defection round") +
  labs(x = "Round",
       y = "Percentage Return",
       color = "Trust Game Phase") +
  theme_bw() +
  theme(legend.position = "bottom",
        legend.box = "vertical", # This will stack the legends vertically
        legend.key = element_blank(),
        legend.margin = margin(t = 0.2, b = 0.2, unit = "pt"),
        legend.box.margin = margin(t = 0, b = 0, unit = "pt"),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10)) +
  scale_color_manual(values = c("darkblue", "darkgreen"),
                     labels = c("Pre-manipulation", "Post-manipulation")) +
  scale_fill_manual(values = c("darkblue", "darkgreen"),
                    guide = "none") + # Hide the fill legend
  guides(color = guide_legend(override.aes = list(fill = NA)))+  # Remove fill for color legend keys
  guides(color = guide_legend(title = "Participant Returns Proportions", override.aes = list(fill = NA)),
         linetype = guide_legend(title = "Defection round")) +
  facet_wrap(~condition.f)

@@

Mixed-effects analysis on the percentage returns shows a significant
main effect of Phase (Pre vs. Post RTG game),
`r papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f`, with
higher percentage returns in the Pre RTG compared to the Post.
Importantly, we also find an interaction between Condition and Phase
(RTG pre- vs. post-manipulation),
`r papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f_condition_f`.
As shown in Figure \@ref(fig:boxPlots), post-hoc tests confirm that,
contrary to our expectations, there was a decrease in the percentage
returned only in the Manipulation condition, pre - post,
`r papaja::apa_print(prepost_bycond)$full_result$Manipulation_Pre_post`,
but no change in the Control condition. We find no interaction between
Phase, Condition and RS, suggesting that there was no difference between
RS groups for this interaction.

There was also a significant main effect of Investment,
`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$inv_scaled)`,
such that higher investments were associated with higher percentage
returns indicating positive reciprocity. An Investment by Condition
interaction,
`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$condition_f_inv_scaled)`,
reflected that returns were more affected by investments in the Control
condition. We also find a three way interaction between Phase,
Investment and RS,
`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f_inv_scaled_high_RS)`,
showing that the differentiated effect of the investment on the
proportion returned by RS group is itself moderated by the Phase (pre-
vs post manipulation). Finally, we find a four-way interaction between
Condition, Phase, Investment and RS
`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f_condition_f_inv_scaled_high_RS)`.

>>>>

---
title: "When Forgiveness Backfires: Rejection Sensitivity and Cooperative Behavior Following Exposure to Adaptive Forgiving Agents"
author:
  "Ismail Guennouni^1,2,3,4^*, Georgia Koppe^1,2,3^$^\\dagger$, Christoph Korn^4^$^\\dagger$"
abstract: |
bibliography: "bib/exposure.bib"
csl: apa.csl
output:
  bookdown::pdf_document2:
   # bookdown::word_document2:
    keep_tex: true
    toc: false
header-includes:
  - \usepackage{lscape}
editor_options: 
  markdown: 
    wrap: 72
---

@@

Can exposure to forgiving partners improve interpersonal cooperation? Attachment theory suggests positive relational experiences can correct negative internal working models, but individuals high in Rejection Sensitivity (RS)—characterized by anxious expectations of rejection—may struggle to benefit from such experiences. We tested this using a randomized experiment (N = 206) in which participants played repeated trust games with HMM-based artificial agents that simulate human-like trust dynamics. After a baseline game, participants were exposed to either forgiving agents (Manipulation) or standard agents (Control), then played a final game with a standard agent. Contrary to predictions, forgiveness exposure *reduced* subsequent cooperation—participants appeared to perceive the standard post-exposure agent as less cooperative by comparison (a negative contrast effect). Critically, RS moderated specific behavioral patterns but not overall cooperation levels: high RS participants failed to recover cooperation after trust violations and became *less* responsive to partner behavior following exposure, whereas low RS participants showed normal recovery and became *more* responsive. These findings suggest that positive relational experiences do not universally promote cooperation, and that high RS individuals may require interventions targeting their capacity to update expectations rather than simply providing positive experiences.

@@

This study found that exposing people to forgiving partners in economic games unexpectedly decreased their subsequent cooperation—likely because normal partners seemed less cooperative by comparison. Individuals high in rejection sensitivity showed a distinct pattern: while they detected trust violations as readily as others, they failed to restore cooperation afterward and became less responsive to their partner's behavior. In contrast, those low in rejection sensitivity appeared to learn from the positive exposure and became more reciprocal. These findings suggest that simply providing positive social experiences may not benefit everyone equally, and that individuals prone to rejection sensitivity may need targeted support to translate positive experiences into lasting behavioral change.

@@

# Helper function to format emmeans contrast results in APA style
format_emmeans_apa <- function(result_row) {
  est <- result_row$estimate
  se <- result_row$SE
  df <- result_row$df
  t_val <- result_row$t.ratio
  p_val <- result_row$p.value

  # Format p-value
  if (p_val < .001) {
    p_str <- "p < .001"
  } else {
    p_str <- paste0("p = ", papaja::printp(p_val))
  }

  # Format the string
  sprintf("$\\Delta M$ = %.2f, $SE$ = %.2f, $t$(%.1f) = %.2f, %s",
          est, se, df, t_val, p_str)
}

@@

One explanation for such social challenges lies in early caregiver
experiences. Attachment theory [@bowlby_attachment_1978] suggests that
the quality of these relationships shapes our capacity for secure
attachments and trust. Individuals with higher levels of insecure
attachment may recall negative trust-related experiences more easily,
report fewer positive trust experiences, and use less constructive
coping strategies when trust is broken [@mikulincer_attachment_1998].
These insecure attachment patterns are often associated with heightened
rejection sensitivity (RS), a tendency to anxiously expect, readily
perceive, and intensely react to rejection [@downey_implications_1996;
@downey_early_1997]. RS has been linked to the development of various
mental health conditions, including depression, anxiety, personality
disorders, and self-harm [@gao_associations_2017]. Individuals high in
RS show attentional biases towards social threat cues, which may
contribute to difficulties in social interactions
[@berenson_rejection_2009]. A recent meta-analysis revealed prosocial
behavior and interpersonal trust as two key processes of interpersonal
functioning that are markedly impaired in PDs and which are likely to
contribute to interpersonal dysfunction in this population
[@hepp_prosociality_2022]. The interaction of RS and trust-based
constructs has been explored, particularly in Borderline Personality
Disorder (BPD). @miano_rejection_2013 and @richetin_emotional_2018 found
that RS mediated the relationship between BPD features and lower trust
appraisal. @abramov_influence_2022-1 found that higher baseline feelings
of rejection in individuals with BPD predict slower trust formation and
less pronounced declines in trust following trust violations during the
trust game. However, the interaction between *reciprocity* and RS hasn't
been studied as extensively, leaving a gap in our understanding of how
these constructs might interplay.

Given that RS may be a manifestation of maladaptive attachment styles,
it is important to explore whether exposure to consistently forgiving
interaction partners could reshape interpersonal expectations and
behaviors. The *corrective experience hypothesis*, rooted in attachment
theory, suggests that new positive relational experiences can modify
internal working models of relationships [@bowlby_secure_1988]. Research
on social learning [@bandura_social_1977] similarly demonstrates that
individuals model the behavior of those around them, and exposure to
cooperative peers promotes cooperative behavior
[@fowler_cooperative_2010]. In the repeated trust game (RTG) paradigm,
cycles of reciprocated trust enhance cooperative behaviors even among
initially distrustful individuals [@king-casas_getting_2005]. This
perspective predicts that exposure to forgiving partners should increase
subsequent cooperation, as participants internalize more positive
expectations about social interactions. Given that high RS individuals
show heightened sensitivity to social cues, they might be particularly
responsive to such corrective experiences.

@@

We pre-screened participants for high or low RS using a validated
questionnaire, then assigned them exclusively to the trustee role in a
series of trust games. After playing a 15-round RTG with a human-like
HMM investor, they were randomly assigned to either a Control or
Manipulation condition. In the Manipulation condition, participants were
exposed over three RTGs to HMM investors designed with a limited
propensity for retaliation. In the Control condition, participants
played three RTGs against the same human-like HMM. After this exposure
phase, all participants played another 15-round RTG with a human-like
HMM investor, similar to the one in the pre-exposure phase.

Based on the corrective experience hypothesis, we predicted that
forgiveness exposure would increase subsequent cooperation, with high RS
individuals potentially benefiting from positive relational experiences
that challenge their negative expectations. We examined both overall
effects of the manipulation and differential responses based on RS, with
particular attention to how participants respond to trust violations
before and after the exposure phase.

@@

The experiment had a 2 (Condition: Manipulation or Control) by 2 (RS :
High or Low) by 2 (Phase: Trust-Game Pre-Exposure, Trust-Game Post-Exposure) design, with repeated measures on the Phase factor (Figure
\@ref(fig:HMMPanels).A). Participants within each pre-screened RS group
were randomly assigned to one of the two levels of the Condition factor,
resulting in 101 participants in the Manipulation condition and 105 in
the Control condition. The games were designed and implemented online
using Empirica v1 [@almaatouq_empirica_2021]. The planned experiment
received approval from the University of Heidelberg's Medical Faculty
ethics commission (ID:S-708/2023) and the experiment was performed in
accordance with the ethics board guidelines and regulations. All
participants provided informed consent prior to their participation.

@@

In order to instigate a potential breakdown of trust, thereby allowing
us to probe efforts to repair it, the computerised agent was programmed
to provide a low investment on round 12 (pre-exposure) and round 13
(post-exposure). On all other rounds, the investor’s actions were
determined by randomly drawing an investment from the state-conditional
distribution, with the state over rounds determined by randomly drawing
the next state from the state-transition distribution as determined from
the net return on the previous round (disregarding the net return
immediately after the pre-programmed low investment rounds).

@@

At the start of the experiment, participants provided informed consent
and were instructed the study would consist of three phases in which
they would face a different other player. Participants were told their
goal was to maximise the number of points in all phases. They were not
told the number of rounds of each phase. Participants were randomly
assigned to either a Control or Manipulation condition. The timeline of
the experiment is shown in Figure \@ref(fig:HMMPanels).A. Phase one
("pre") consisted of a 15 round RTG in which participants took the role
of trustee, facing the same investor over all 15 rounds. On each round,
after being informed about the amount sent by the investor participants
decided how much of the tripled investment to return to the investor,
before continuing to the next round. Phase 2 ("exposure") consisted of
three 7-round RTGs. Participants in the Manipulation condition faced the
forgiving HMM investor and rated the agent on the same attributes as in
the pre-exposure phase. Those in the Control condition faced the
same human-like HMM agent as in the "pre" phase and rated each co-player
on the same attributes. To keep the experience similar to the "pre"
phase, the agent in the Control condition was also designed to send a
very low investment in round 5 of each of the three games. In the
post-exposure phase ("post"), participants in both conditions faced
the same human-like HMM as in "pre" phase.

@@

````{=tex}
\begin{landscape}
```{r HMMPanels, fig.cap= "A: Experiment timeline. Participants (trustees) played RTGs with HMM investor agents. The investor sends investments (multiplied by 3) and participants decide returns. Conditions differ in exposure phase agents. B-D: The artificial investor is a three-state HMM fitted to human data. B: Investment distributions by latent state. C: Transition probabilities to states at t+1 as a function of net return at t; each panel shows a different starting state. D: Forgiving HMM transitions from low-trust state---unlike the human-like HMM, it always exits low-trust and favors high-trust transitions.", include=T, out.width='0.95\\linewidth', fig.align='center'}

@@

# Calculating the width and height for each image (with shorter caption, can use more space)
width_per_image <- ((10.5 - 0.1) / 2) * dpi
height_per_image <- ((5.8 - 0.1) / 2) * dpi
# The geometry string for each image, including spacing (assuming no borders)
geometry_string <- paste0(width_per_image, 'x', height_per_image, '+', spacing, '+', spacing)

@@

We analyzed participants' behavior in the RTG using linear mixed-effects models. First, to examine the effect of the manipulation, we modeled the percentage return (percentage of tripled investment returned to investor) as a function of Phase (RTG game pre vs. post-exposure), Condition (Manipulation vs. Control), Investment, and RS (High vs Low RS group), including all interactions as fixed effects. This model included player-wise random intercepts and slopes for Phase. Second, we analyzed behavior during the Exposure phase specifically, modeling returns with Condition, Investment, and RS and their interactions as fixed effects, along with player-wise random intercepts. Third, to verify the consistency of the HMM agent, we modeled the investments sent by the computerized agent using Condition, Phase, and RS and their interactions as fixed effects. To isolate effects occurring prior to any pre-programmed low investment, we also analyzed returns in rounds preceding the low investment trials only (rounds 1-11 in the pre-exposure phase and rounds 1-12 in the post-exposure phase) using the same model specification. Finally, to rigorously assess participants' reactions to and recovery from the specific instance of pre-programmed low investment, we conducted an event study analysis centered on the low investment round ($t=0$). We analyzed percentage returns in a three-round window ($t-1$ to $t+1$) using a linear mixed-effects model with Phase, Condition, Time Point, and RS Group as fixed effects. We specifically examined two key behavioral responses: the Drop (change in return from $t-1$ to the low investment round) and the Recovery (change in return from the low investment round to $t+1$). The full specification of all statistical models can be found in the supplement.

All models were estimated using the `afex` package [@singmann_afex_2022] in R. We determined the random effects structure by starting with the maximal model and simplifying until convergence was achieved, ensuring the optimal structure [@matuschek_balancing_2017]. A similar process was applied to the models analyzing HMM agent investments and participant ratings. We report differences in marginal means rather than effect sizes, as there is no consensus on effect size calculation for mixed models. $F$-tests used the Kenward-Roger approximation for degrees of freedom. The Investment variable was Z-transformed to facilitate the interpretation of main effects in the presence of interactions. Significant interactions were probed using planned contrasts with the `emmeans` package. We applied the "Sidak" correction for multiple comparisons to control the familywise error rate while maintaining statistical power.

@@

```{r}
# average investment for first 10 rounds pre-exposure
avg_pct <- avg_ret_df %>% filter(roundNum < 11 , gameNum.f == "pre") %>% summarise(avg_inv = mean(inv_pct), avg_ret = mean(ret_pct_na, na.rm=TRUE))
avg_pct
```

@@

```{r gamesPlot, include=T, echo=FALSE, fig.cap="Averages and standard errors of the trustee's return as a percentage of the multiplied investment received (y-axis) by Condition, Phase, and game round (x-axis) averaged across RS groups. The blue line shows the returns in the Pre phase and the green line those in the Post phase. The left Panel shows returns in the Control condition and the right one those in the Manipulation condition. The dotted lines identify the rounds where the pre-programmed one-off low investment occurs. We note lower average returns post vs pre in the Manipulation condition, whilst returns in the Control condition are similar between the two phases.",fig.align="center", fig.width=5, fig.height = 3.3}

# Create a data frame for the vertical lines
vline_data <- data.frame(
  xintercept = c(12, 13),
  Low_inv_round = factor(c("Pre-phase", "Post-phase"), levels = c("Pre-phase", "Post-phase"))
)

# Plot
ggplot(avg_ret_df %>% filter(gameNum.f %in% c("pre","post")), aes(x=as.factor(roundNum), y=ret_pct_na, group=gameNum.f, color = gameNum.f, fill=gameNum.f)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun.data = mean_se, geom = "ribbon", aes(ymin=..ymin.., ymax=..ymax..),
               alpha = 0.3, linetype = 0) +
  geom_vline(data = vline_data, aes(xintercept = xintercept, linetype = Low_inv_round), color = "black", linewidth = 0.6) +
  scale_linetype_manual(values = c("Pre-phase" = "dotted", "Post-phase" = "dashed"),name = "Low investment round") +
  labs(x = "Round",
       y = "Percentage Return",
       color = "Trust Game Phase") +
  theme_bw() +
  theme(legend.position = "bottom",
        legend.box = "vertical", # This will stack the legends vertically
        legend.key = element_blank(),
        legend.margin = margin(t = 0.2, b = 0.2, unit = "pt"),
        legend.box.margin = margin(t = 0, b = 0, unit = "pt"),
        legend.title = element_text(size = 9),
        legend.text = element_text(size = 8)) +
  scale_color_manual(values = c("darkblue", "darkgreen"),
                     labels = c("Pre-phase", "Post-phase")) +
  scale_fill_manual(values = c("darkblue", "darkgreen"),
                    guide = "none") + # Hide the fill legend
  guides(color = guide_legend(override.aes = list(fill = NA)))+  # Remove fill for color legend keys
  guides(color = guide_legend(title = "Participant Returns Proportions", override.aes = list(fill = NA)),
         linetype = guide_legend(title = "Low investment round")) +
  facet_wrap(~condition.f)

@@

Participants returned higher percentages in the Pre phase compared to
the Post phase
(`r papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f`). This
effect was moderated by Condition
(`r papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f_condition_f`):
contrary to our expectations, participants in the Manipulation condition
decreased their returns from pre to post
(`r papaja::apa_print(prepost_bycond)$full_result$Manipulation_Pre_post`),
while those in the Control condition showed no change (Figure
\@ref(fig:boxPlots)). RS did not moderate this Condition × Phase
interaction.

Higher investments elicited higher percentage returns, indicating
positive reciprocity
(`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$inv_scaled)`).
This relationship was stronger in the Control condition than in the
Manipulation condition
(`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$condition_f_inv_scaled)`).
The effect of investment on returns varied by RS group and Phase
(`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f_inv_scaled_high_RS)`),
and a four-way interaction indicated that these patterns further
differed across Conditions
(`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f_condition_f_inv_scaled_high_RS)`).

# Define colors for Phase (different from event study's Condition colors)
phase_cols <- c("pre" = "#E69F00", "post" = "#0072B2")

# Generate the plot data from afex_plot
ret_all_data <- afex_plot(mod_returns_pct_RS, x = "condition.f", trace = "phase.f", dodge = 0.8, error = "within",
                     mapping = c("linetype", "shape", "fill"),
                     factor_levels = list(condition.f = c("Control", "Manipulation")),
                     legend_title = "Phase",
                     emmeans_arg = list(p.adjust.method = "tukey"))

# Extract the data used in afex_plot
plot_data_ret <- ret_all_data$data

# Generate the bar chart for returns
ret_all <- ggplot(plot_data_ret, aes(x = condition.f, y = y, fill = phase.f)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, color = "black", size = 0.3) +
  geom_errorbar(aes(ymin = y - 1.96 * SE, ymax = y + 1.96 * SE),
                position = position_dodge(width = 0.7), width = 0.2, size = 0.5) +
  scale_fill_manual(values = phase_cols, labels = c("Pre-phase", "Post-phase")) +
  theme_bw(base_size = 11) +
  labs(y = "% Returns", x = "Condition", fill = "Trust Game Phase") +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11)
  ) +
  coord_cartesian(ylim = c(0, 0.6)) +
  # Significance bracket for Manipulation pre vs post (moved higher to avoid overlap)
  geom_segment(aes(x = 1.825, xend = 2.175, y = 0.54, yend = 0.54), inherit.aes = FALSE, size = 0.5) +
  geom_segment(aes(x = 1.825, xend = 1.825, y = 0.54, yend = 0.52), inherit.aes = FALSE, size = 0.5) +
  geom_segment(aes(x = 2.175, xend = 2.175, y = 0.54, yend = 0.52), inherit.aes = FALSE, size = 0.5) +
  geom_text(aes(x = 2, y = 0.56, label = "**"), inherit.aes = FALSE, size = 5)

# Generate the plot data from afex_plot for investments
inv_all_data <- afex_plot(mod_invs_RS, x = "condition.f", trace = "phase.f", dodge = 0.8, error = "within",
                     mapping = c("linetype", "shape", "fill"),
                     factor_levels = list(condition.f = c("Control", "Manipulation")),
                     legend_title = "Phase")

# Extract the data used in afex_plot
plot_data_inv <- inv_all_data$data
